\myTitle

### 2024-02-03 ms.Rnw
updated to match arxiv submission

%\documentclass[12pt]{article}
\documentclass[11pt]{article} \usepackage{fullpage}
\pdfoutput=1 
\newcommand\arxiv[2]{#1}  %% for arxiv version
%% \newcommand\arxiv[2]{#2}  %% suppressing arxiv formatting

<<setup,echo=F,message=F,cache=F>>=
# computational intensity, i, also known as run level
i <- 3
library(metapoppkg)
Ncities <- ncol(metapoppkg::incidence) - 1 # first column is date
U <- switch(i,20,Ncities,Ncities)
library(foreach)
library(doParallel)
library(doRNG)
cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
if(is.na(cores)) cores <- detectCores()
registerDoParallel(cores)

R0_M6 <- function (params, be = T) 
{
    a = params[ifelse(be, "alpha_be1", "alpha_af1")]
    b = params[ifelse(be, "Beta_be1", "Beta_af1")]
    D = params["D1"]
    mu = params[ifelse(be, "mu_be1", "mu_af1")]
    unname((a + (1 - a) * mu) * b * D)
}

ci_c <- readRDS("ci_c.rds")  # generated by si/si.Rnw
ci_uc <- readRDS("ci_uc.rds") # generated by si/si.Rnw
@

\newcommand\LiMobility{M$_1$}
\newcommand\LiParams{M$_2$}
\newcommand\BenchmarkIID{M$_3$}
\newcommand\BenchmarkAR{M$_4$}
\newcommand\RevisedModelUnconstrained{M$_5$}
\newcommand\RevisedModelConstrained{M$_6$}

\newcommand\Rzero{\mathcal{R}_0}
\newcommand\before{\mathrm{be}}
\newcommand\after{\mathrm{af}}
\newcommand\code[1]{\texttt{#1}}

\newcommand\suppSecEnkf{S4}
\newcommand\suppSecOurModel{S1}
\newcommand\suppSecMobility{S1.2}
\newcommand\suppSecLiModel{S1.3}
\newcommand\suppSecBenchmark{S2}
\newcommand\suppSecIBPF{S3}
\newcommand\suppSecEnKF{S4}
\newcommand\suppSecUnconstrained{S5} %% profiles for the unconstrained li23 model
\newcommand\suppSecInference{S6} %% now profile likelihood CIs
\newcommand\suppSecResiduals{S7} %% now anomaly analysis
\newcommand\suppSecMetapoppkg{S8}

\newcommand\suppFigSim{S2}
\newcommand\suppTableResults{S1}
\newcommand\suppFigFilterComparison{S9}

\newcommand\stdError[1]{}
%\newcommand\stdError[1]{{\small(#1)}}

\newcommand\includeRates[1]{}

\usepackage[sort&compress,numbers]{natbib}
\usepackage{times}
\usepackage{url}
\usepackage{enumitem}
\usepackage[hidelinks,colorlinks=true,linkcolor=blue,citecolor=blue]{hyperref}
\usepackage{tikz}

\usepackage{xcolor}
%orange for EI
\definecolor{orange}{rgb}{1,0.5,0}
\newcommand\ei[2]{\sout{#1} \textcolor{orange}{#2}}
\newcommand\eic[1]{\textcolor{orange}{[#1]}}
%green for PN
\definecolor{green}{rgb}{0,0.5,0}
\newcommand\pn[2]{\sout{#1} \textcolor{green}{#2}}
\newcommand\pnc[1]{\textcolor{green}{[#1]}}
%purple for JF
\definecolor{purple}{rgb}{0.5,0,1}
\newcommand\jf[2]{\sout{#1} \textcolor{purple}{#2}}
\newcommand\jfc[1]{\textcolor{purple}{[#1]}}
%teal for TODO
\definecolor{teal}{rgb}{0,0.5,0.8}
\newcommand\todo[1]{\textcolor{teal}{[TO DO? #1]}}

\newenvironment{sciabstract}{%
\begin{quote} \bf}
{\end{quote}}

\title{Machine Learning for Mechanistic Models of Metapopulation Dynamics}

@@

\begin{abstract}
Mathematical models in ecology and epidemiology must be consistent with observed data in order to generate reliable knowledge and evidence-based policy. Metapopulation systems, which consist of a collection of sub-populations at various locations, pose technical challenges in statistical inference due to nonlinear, stochastic interactions. Difficulties encountered in these methodological tasks can obstruct the core scientific questions concerning the link between the mathematical models and the data. Progress in statistically efficient simulation-based inference for partially observed stochastic dynamic systems has enabled the development of statistically rigorous approaches to the analysis of nonlinear but low-dimensional systems. Recently, an algorithm has been developed which enables comparable inference for higher-dimensional models arising in metapopulation systems. The COVID-19 pandemic provides a situation where mathematical models and their policy implications were widely visible, and we revisit an influential metapopulation model used to inform basic epidemiological understanding early in the pandemic. Our methods support self-critical data analysis, enabling us to identify and address model limitations, and leading to a new model with substantially improved statistical fit and parameter identifiability. Our results suggest that the lockdown initiated on January 23, 2020 in China was more effective than previously thought. We proceed to recommend statistical analysis standards for future metapopulation system modeling.
\end{abstract}

@@

Biological populations may be structured into a collection of densely-populated communities separated by sparsely populated regions.
The communities, which may be cities in a human context, comprise a metapopulation.
Motivation for metapopulation modeling arises when some essential feature of the population dynamics cannot be understood from looking at a single location.
Dynamics of persistence through local extinctions and reintroductions have been extensively studied in ecology \cite{hanski98,mackenzie09}.
In epidemiology, metapopulation dynamics can be a barrier to the regional elimination and eventual eradication of a pathogen, and may determine the successful invasion of a new pathogen or a new strain of an existing pathogen \cite{metcalf21}.
In other situations, spatiotemporal dynamics may be an unavoidable component of the system under study without being the focus of the investigation \cite{zhang22,wheeler23}.

A recent growth in the study of metapopulation dynamics has been driven partly by the COVID-19 pandemic \cite{li20,wu20,wang22,prieto22,cascante-vega22,pizzuti20,alleman21,yang21,engebretsen23} and in part by methodological advances facilitating the fitting of metapopulation models to spatiotemporal data.
Until the start of this millenium, developing dynamic models with both statistical and scientific justification was a longstanding open problem for even a single community \cite{bjornstad01}.
Over the past two decades, new algorithms \cite{ionides06,toni09,andrieu10,ionides15} and software \cite{king16,kristensen16,devalpine17}, together with ever-increasing computational resources, have enabled routine inference for low-dimensional nonlinear partially observed stochastic dynamic systems.
However, fundamental algorithmic scalability issues known as the ``curse of dimensionality'' lead to difficulties with the high-dimensional systems arising in metapopulation inference.
These issues are clearest for Monte Carlo techniques based on importance sampling \cite{bengtsson08} but are also evident in the need for variational approximations for large Monte Carlo Markov Chain (MCMC) calculations \cite{blei17}. 
The quest for statistical methods suitable for metapopulation models of scientific interest has motivated the development of various methodological approaches.
Nevertheless, limitations have persisted in the capability to carry out flexible model-based inference and rigorous model criticism.
Thus, data analysis for metapopulation models has lagged behind low-dimensional time series analysis of biological systems.
Recent developments enable this gap to be closed, as we demonstrate via a reanalysis of COVID-19, viewed from the context of the ability to drawm evidence-based scientific conclusions about the dynamics of the emerging pandemic in January and February 2020.

\section*{Review of metapopulation models and inference methods}
\label{sec:review}

Biological systems are characterized by nonlinear stochastic dynamics together with incomplete and noisy measurements  \cite{bjornstad01}.
We therefore focus on the class of partially observed Markov process (POMP) models \cite{breto09}, acknowledging that deterministic models can be conceptually useful but are problematic as statistical explanations of noisy systems \cite{king15,wheeler23}.
The Markov property asserts that the dynamic process has no memory conditional on its current state, which is algorithmically convenient while being scientifically nonrestrictive since we can choose what to include in the state.
Metapopulation models consider a multivariate system state at each location and so we require methods tailored for high-dimensional POMP models.
Simplifications arise if models and data are limited to binary presence-absence, or a small discrete set of values at each location \cite{mackenzie09}, but we focus on situations where abundance data are available, such as case reports for infectious diseases.

A natural place to look for statistical methodology applicable to metapopulation models is among the techniques developed for inferring population dynamics at a single location, reviewed by \cite{funk20,auger-methe21}.
Commonly implemented approaches for POMP models can be categorized as  (i) variants of MCMC; (ii) matching summary statistics between data and simulations; (iii)  linearization;   (iv) particle filters (i.e., sequential Monte Carlo).

In principle, MCMC techniques enable Bayesian inference or maximum likelihood via expectation-maximization algorithms \cite{cappe05}.
In practice, successful MCMC for metapopulation models requires careful model-specific algorithm development \cite{whitehouse23}.

Matching summary statistics of simulations to the corresponding data statistic is, in principle, a readily applicable inference approach for a wide class of models including metapopulation models.
In the context of Bayesian inference this is called Approximate Bayesian Computing \cite{conlan12}.
However, informative, low-dimensional summary statistics can be hard to construct even for low-dimensional nonlinear systems.
This can make summary statistic methods statistically inefficient \cite{fasiolo16}.

Population dynamics may be approximately linear on a log scale, and this has been used to develop linearization methods for epidemiological time series analysis \cite{bjornstad02} that have been extended to metapopulation analysis \cite{xia04}.
This provides a numerically convenient set of tools, but requires scientists to work within a limited class of models.

The ensemble Kalman filter (EnKF) developed in the context of massive geophysical model, combines an ensemble representation with a computationally efficient update rule inspired by the scalable linear Kalman filter, providing an approach with excellent scalability \cite{evensen09book,evensen22}.
For biological systems, EnKF was first demonstrated as a computationally convenient tool for compartment models at a single location \cite{shaman12,yang14}.
Subsequently, it has been applied for epidemiological metapopulation inference \cite{li20,kramer20-plos-cb}.
However, the linearization in the EnKF filter update rule can be problematic for highly nonlinear systems \cite{evensen22}.
Further, a linear update rule is not appropriate for small, discrete populations unless EnKF is embedded within a MCMC algorithm \cite{katzfuss19}.

For low-dimensional systems, particle filter methods are applicable to a flexible class of models: they permit consideration of arbitrary nonlinear dynamics and require the model to be specified only via a simulator \cite{breto09,king16}.
Particle filters enable statistically efficient use of data, since they provide an evaluation of the likelihood function required for Bayesian or likelihood-based inference, with approximation resulting only from finite Monte Carlo effort.
For high-dimensional systems, scalability considerations demand further approximations since particle filters suffer acutely from the ``curse of dimensionality'' \cite{bengtsson08}.
In contrast, block particle filter (BPF) methods achieve scalability by updating particles through localized resampling, rather than employing the linear update of EnKF \cite{rebeschini15}.
It is an empirical question which of these approximations is more successful on metapopulation models,  with prior evidence favoring BPF \cite{ionides23-jasa}.

Fitting complex models to large datasets using computationally intensive methods is the domain of machine learning.
Some machine learning models, such as artificial neural networks, lack a mechanistic interpretation.
Here, we focus instead on machine learning for mechanistic models.
We not only compare available machine learning methods for fitting metapopulation models but, importantly, we also consider issues of model criticism.
Specifically, we introduce a model diagnosis procedure to mitigate the risk of selecting an inappropriate model, an essential step in informing public policy decisions.

\section*{Metapopulation analysis of COVID-19 spread in China}
\label{sec:covid}
Models are an irreplaceable tool to inform public policy, despite delicate issues in their implementation and interpretation \cite{mccabe21,saltelli20}.
Some of the difficulties are operational, others are conceptual.
Operationally, we seek to fit complex models using statistically valid, reproducible and transparent methods.
Conceptual difficulties arise when drawing causal conclusions from fitting models to observational data, giving rise to opportunities for incorrect conclusions due to missing variables or other forms of model misspecification.
A strength of metapopulation modeling is that the model can be build upon established scientific knowledge, which may give some protection against gross forms of model misspecification (for example, asserting the existence of latent dynamic variables which simply do not exist).
A model assimilated to data guarantees that assumptions have been framed in a way consistent with certain facts, and evidence for predictive skill can support the value of the model construction.
Indeed, it can be practically impossible to make sense of the nonlinear stochastic interactions driving biological dynamics without representing them via a model \cite{mccabe21,lofgren14}.
We therefore seek to harness the power of models while assessing, acknowledging and minimizing their weaknesses.

As an example, we reconsider the influential analysis of COVID-19 from early in the pandemic by Li et al. \cite{li20}.
This analysis provided estimates of transmission parameters and the effect of the lockdown in China using the limited data available at the time.
Other teams have fitted models to address similar questions \cite{kraemer20,yang21,brett23} but the study by \cite{li20} is distinctive for fitting a stochastic mechanistic metapopulation model to extensive spatiotemporal data.
The results were published in May, 2020, based on reported cases from January 10 to February 8 of that year.
The state-of-the-art spatiotemporal analysis was possible on an urgent timescale because the team of researchers had developed their methodology in a sequence of previous situations \cite{shaman12,yang14,yang15,pei18}.
The paper is written with attention to reproducibility, and the main results are strengthened by various supporting analyses in an extensive supplement.
While examining the points mentioned above, we have identified various limitations that could have been mitigated by adhering to the aforementioned recommendations.
Our goal is not to criticize any specific paper, but rather to build on the timely analysis of \cite{li20} to demonstrate how recently developed techniques provide possibilities to carry out improved data analysis in future.

@@

\begin{figure}
\begin{center}
%%%%% SEAIR diagram
\resizebox{11cm}{!}{
\begin{tikzpicture}[
  square/.style={rectangle, draw=black, minimum width=0.8cm, minimum height=0.8cm, rounded corners=.1cm, fill=blue!8},
  travel/.style={circle, draw=black, minimum width=0.85cm, minimum height=0.8cm, fill=green!8},
  report/.style={shape=regular polygon, regular polygon sides=8, draw, fill=red!8,minimum size=0.9cm,inner sep=0cm},
  bendy/.style={bend left=25},
  >/.style={shorten >=0.25mm}, % redefine arrow to stop short of node
  >/.tip={Stealth[length=1.5mm,width=1.5mm]} % redefine arrow style
]
\tikzset{>={}}; % this is needed to implement the arrow redefinition

@@

For our metapopulation system, the sub-populations are 373 provincial cities in China (meaning cities with administrative responsibility for an entire region) and the data are daily reported COVID-19 cases. 
COVID-19 dynamics are represented by an Susceptible-Exposed-Asymptomatic-Infectious-Recovered (SEAIR) epidemic model. 
Questions of urgent interest early in the pandemic include the relative transmissibility of reported to unreported cases, the fraction of unreported cases, and the effect on transmission of movement restrictions imposed on and around January 23 \cite{li20}.
The model structure is illustrated by the flow diagram in Fig.~\ref{fig:flow_diagram}.
We consider different model implementations within this structure, with full model specifications provided by the equations and parameter values in Supplementary Sec.~\suppSecOurModel.

Our starting point is model {\LiMobility} which is based on the model of \cite{li20} and is fully described in Supplementary Sec.~\suppSecLiModel.
We consider the full dataset, from January 10 to February 8, with transmission parameters re-estimated following the lockdown on January 23; these correspond to the periods~1 (January 10 to January 22) and 3 (January 24 to February 8) of \cite{li20}.
Some minor differences between {\LiMobility} and \citep{li20} were introduced to enable us to place their model within the general metapopulation framework of spatiotemporal partially observed  continuous-time Markov processs described by \citep{ionides23-jasa}.
Despite these modifications, simulations from {\LiMobility}, using the parameters of \citep{li20}, closely match simulations from the code provided by \cite{li20} (Supplementary Sec.~\suppSecLiModel).
However, inspection of the mobility data reveals that some small cities have no recorded incoming travelers, and therefore no possibility of a SARS-CoV2 introduction within {\LiMobility} (or the model of \cite{li20}) (Supplementary Sec.~\suppSecMobility).
This minor limitation formally results in a likelihood of zero for {\LiMobility} (i.e., it is impossible for the simulation model to reproduce the observed spatiotemporal dataset), and hence a log-likelihood of $-\infty$.

@@

We addressed the problematic mobility data in {\LiMobility} by adding some additional transportation based on a gravity movement model, as described in Supplementary Sec.~\suppSecMobility, giving rise to model~{\LiParams}.
We implemented an additional adjustment between models {\LiMobility} and {\LiParams} to align the measurement model with the ensemble Kalman filter (EnKF) inference method presented by \cite{li20}.
That EnKF implementation involved specifying a quantity called the observation error variance, defined as a function of the observed cases, to quantify the uncertainty in the measurements.
Within the POMP specification, the measurement variance can depend on the latent state but not directly on the observed data.
To interpret the choice of EnKF observation variance within the POMP framework, we specified the measurement model for~{\LiParams} to have equivalent scaling to the choice of \citep{li20}, but with dependence on the reported cases replaced by dependence on the modeled, but unobserved, exact case count.

@@

Table~1 shows that model~{\RevisedModelUnconstrained} outperforms simple statistical benchmarks, obtaining a competitive likelihood with relatively few parameters.
From a statistical perspective, {\RevisedModelUnconstrained} is therefore an adequate statistical description of the data.
However, some parameters of {\RevisedModelUnconstrained} were weakly identified by the data, especially in the pre-lockdown time interval within which there were relatively few reported cases (Supplementary Sec.~\suppSecUnconstrained).
When the evidence about the model parameters in the data is weak, the ambiguity may be resolved by other, unmodeled and poorly understood, aspects of the data.
This risks leading to undesirable situations where substantial conclusions about questions of interest could be driven by the weaknesses of the model rather than its strengths.
In Supplementary Sec.~\suppSecUnconstrained, we show how the flexibility of {\RevisedModelUnconstrained} can be used to obtain a high likelihood via an unplausibly long estimated duration of infection during the pre-lockdown period, with the estimate suddenly reducing after lockdown.
 We resolved this issue by constraining the latent and infectious periods to be the same before and after lockdown, leading to model {\RevisedModelConstrained}.
 The additional constraints of {\RevisedModelConstrained} lead to a small loss of likelihood compared to {\RevisedModelUnconstrained}, but the fit remains competitive compared to the benchmark models, and the stronger identifiability facilitates the interpretation of estimated parameters.
Calculating the log-likelihood for each model in Table~\ref{tab:loglik} requires extensive computation to produce a single number which contains essentially all the information about the statistical fit of the model.
However, additional work is required to understand what characteristics of the models and data causes the differences in these numbers, and the practical consequences of the numerical results.

>>>>

%\documentclass[12pt]{article}
\documentclass[11pt]{article} \usepackage{fullpage}
\pdfoutput=1

\input{header}
\newcommand\stdError[1]{}
%\newcommand\stdError[1]{{\small(#1)}} 
\newcommand\includeRates[1]{}

%% \newcommand\arxiv[2]{#1}  %% for arxiv version
\newcommand\arxiv[2]{#2}  %% suppressing arxiv formatting

<<setup,echo=F,message=F,cache=F>>=
# computational intensity, i, also known as run level
i <- 3
library(metapoppkg)
library(foreach)
library(doParallel)
library(doRNG)
library(tidyverse)
library(patchwork)

Ncities <- ncol(metapoppkg::incidence) - 1 # first column is date
U <- switch(i,20,Ncities,Ncities)
cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
if(is.na(cores)) cores <- detectCores()
registerDoParallel(cores)

ci_c <- readRDS("ci_c.rds")  # generated by si/si.Rnw
ci_uc <- readRDS("ci_uc.rds") # generated by si/si.Rnw

library(knitr)
opts_knit$set(concordance=TRUE)
opts_chunk$set(
  # cache=FALSE,
  cache=TRUE,   ## caution: use cache=TRUE only when making text edits
  progress=TRUE,
  prompt=TRUE,
  highlight=FALSE,
  tidy=TRUE,
  tidy.opts=list(
    keep.blank.line=FALSE
  ),
  comment="",
#  warning=FALSE,
#  message=FALSE,
  error=TRUE,
  echo=FALSE,
  strip.white=TRUE,
  results="markup",
  fig.path="figure/",
  fig.lp="fig:",
  fig.align="center",
  fig.show="asis",
  dev="png",
  dpi=300,
  dev.args=list(
    bg="transparent",
    pointsize=9
  )
)

@

\usepackage[sort&compress,numbers]{natbib}

\title{\myTitle}

@@

\begin{abstract}
Mathematical models in ecology and epidemiology must be consistent with observed data in order to generate reliable knowledge and evidence-based policy. Metapopulation systems, which consist of a network of connected sub-populations, pose technical challenges in statistical inference due to nonlinear, stochastic interactions. Numerical difficulties encountered in conducting inference can obstruct the core scientific questions concerning the link between the mathematical models and the data. Recently, an algorithm has been developed which enables effective likelihood-based inference for the high-dimensional partially observed stochastic dynamic models arising in metapopulation systems. The COVID-19 pandemic provides a situation where mathematical models and their policy implications were widely visible, and we use the new inferential technology to revisit an influential metapopulation model used to inform basic epidemiological understanding early in the pandemic. Our methods support self-critical data analysis, enabling us to identify and address model limitations, and leading to a new model with substantially improved statistical fit and parameter identifiability. Our results suggest that the lockdown initiated on January 23, 2020 in China was more effective than previously thought. We proceed to recommend statistical analysis standards for future metapopulation system modeling.
\end{abstract}

@@

Biological populations may be structured into a collection of densely-populated communities separated by sparsely populated regions.
The network of communities, which may be cities in a human context, comprise a metapopulation.
Motivation for metapopulation modeling arises when some essential feature of the population dynamics cannot be understood from looking at a single location.
Dynamics of persistence through local extinctions and reintroductions have been extensively studied in ecology \cite{hanski98,mackenzie09}.
In epidemiology, metapopulation dynamics can be a barrier to the regional elimination and eventual eradication of a pathogen, and may determine the successful invasion of a new pathogen or a new strain of an existing pathogen \cite{metcalf21}.
In other situations, spatiotemporal dynamics may be an unavoidable component of the system under study without being the focus of the investigation \cite{zhang22,wheeler23}.

Mathematical models for biological systems are also used to inform public policy, despite delicate issues in their implementation and interpretation \cite{saltelli20}.
Indeed, it can be practically impossible to make sense of the nonlinear stochastic interactions driving biological dynamics without representing them via a model \cite{mccabe21,lofgren14}.
However. both operational and conceptual difficulties arise when developing these models.
Operationally, we seek to fit complex models using statistically valid, reproducible and transparent methods.
Conceptual difficulties arise when drawing causal conclusions from fitting models to observational data, giving rise to opportunities for incorrect conclusions due to missing variables or other forms of model misspecification.
A model assimilated to data guarantees that assumptions have been framed in a way consistent with certain facts, and evidence for predictive skill can support the value of the model construction.

A recent growth in the study of metapopulation dynamics has been driven partly by the COVID-19 pandemic \cite{li20,wu20,wang22,prieto22,cascante-vega22,pizzuti20,alleman21,yang21,engebretsen23} and in part by methodological advances facilitating the fitting of metapopulation models to spatiotemporal data.
Until the start of this millenium, developing dynamic models with both statistical and scientific justification was a longstanding open problem for even a single community \cite{bjornstad01}.
Over the past two decades, new algorithms \cite{ionides06,toni09,andrieu10,ionides15} and software \cite{king16,kristensen16,devalpine17}, together with ever-increasing computational resources, have enabled routine inference for low-dimensional nonlinear partially observed stochastic dynamic systems.
However, fundamental algorithmic scalability issues known as the ``curse of dimensionality'' lead to difficulties with the high-dimensional systems arising in metapopulation inference.
These issues are clearest for Monte Carlo techniques based on importance sampling \cite{bengtsson08} but are also evident in the need for variational approximations for large Monte Carlo Markov Chain (MCMC) calculations \cite{blei17}. 
Thus, data analysis for metapopulation models has lagged behind the analysis of low-dimensional time series data for biological dynamics.
Recent developments enable this gap to be closed, as we demonstrate via a reanalysis of COVID-19, viewed from the context of the ability to drawm evidence-based scientific conclusions about the dynamics of the emerging pandemic in January and February 2020.

Biological systems are characterized by nonlinear stochastic dynamics together with incomplete and noisy measurements  \cite{bjornstad01}.
We therefore focus on the class of partially observed Markov process (POMP) models \cite{breto09}, acknowledging that deterministic models can be conceptually useful but are problematic as statistical explanations of noisy systems \cite{king15,wheeler23}.
The Markov property asserts that the dynamic process has no memory conditional on its current state, which is algorithmically convenient while being scientifically nonrestrictive since we can choose what to include in the state.
Metapopulation models consider a multivariate system state at each location and so we require methods tailored for high-dimensional POMP models.
Simplifications arise if models and data are limited to binary presence-absence, or a small discrete set of values at each location \cite{mackenzie09}, but we are concerned with situations where time series of abundance data are available, such as case reports for infectious diseases.
We focus on two inferential approaches for high-dimensional POMP models, the block particle filter (BPF) and the ensemble Kalman filter (EnKF).
Other alternatives are reviewed in Supplementary Sec.~{\suppSecReview}.

EnKF was developed in the context of massive geophysical models.
It combines an ensemble representation of the latent state with a computationally efficient update rule inspired by the scalable linear Kalman filter, providing an approach with excellent scalability \cite{evensen09book,evensen22}.
For biological systems, EnKF was first demonstrated as a computationally convenient tool for compartment models at a single location \cite{shaman12,yang14}.
Subsequently, it has been applied for epidemiological metapopulation inference \cite{li20,kramer20-plos-cb}.
However, the linearization in the EnKF filter update rule can be problematic for highly nonlinear systems \cite{evensen22,ionides23-jasa}.
Further, a linear update rule is not appropriate for small, discrete populations unless EnKF is embedded within a MCMC algorithm \cite{katzfuss19}.
By contrast, particle filter methods \cite{doucet11} avoid linearization and are directly applicable to discrete and continuous latent states.

For low-dimensional systems, particle filter methods are broadly applicable; they permit consideration of arbitrary nonlinear dynamics and require the model to be specified only via a simulator \cite{breto09,king16}.
Particle filters enable statistically efficient use of data, since they provide an evaluation of the likelihood function required for Bayesian or likelihood-based inference, with approximation resulting only from finite Monte Carlo effort.
For high-dimensional systems, scalability considerations demand further approximations since particle filters suffer acutely from the ``curse of dimensionality'' \cite{bengtsson08}.
The BPF algorithm modifies the particle filter to achieve scalability by carrying out local resampling on spatial neighborhoods known as blocks.
This avoids the linear update rule used by EnKF \cite{rebeschini15}.
It is an empirical question whether the different approximations inherent in EnKF and BPF are successful on metapopulation models, with prior evidence favoring BPF \cite{ionides23-jasa}.
In the following example, we demonstrate that BPF can be effective for a practical metapopulation data analysis.
We show that the resulting likelihood-based inference framework provides opportunities for model criticism, leading to rigorous assessment of model fit and improved advice on public policy decisions.

\section*{Metapopulation analysis of COVID-19 spread in China}
\label{sec:covid}

We reconsider the influential analysis of COVID-19 from early in the pandemic by Li et al. \cite{li20}.
This analysis provided estimates of transmission parameters and the effect of the lockdown in China using the limited data available at the time.
Other teams have fitted models to address similar questions \cite{kraemer20,yang21,brett23} but the study by \cite{li20} is distinctive for fitting a stochastic mechanistic metapopulation model to extensive spatiotemporal data.
The results were published in May, 2020, based on reported cases from January 10 to February 8 of that year.
The state-of-the-art spatiotemporal analysis was possible on an urgent timescale because the team of researchers had developed their methodology in a sequence of previous situations \cite{shaman12,yang14,yang15,pei18}.
The paper is written with attention to reproducibility, and the main results are strengthened by various supporting analyses in an extensive supplement.
While examining the points mentioned above, we have identified various limitations that could have been mitigated by adhering to the aforementioned recommendations.
Our goal is not to criticize any specific paper, but rather to build on the timely analysis of \cite{li20} to demonstrate how recently developed techniques provide possibilities to carry out improved data analysis in future.

@@

\begin{figure}
\begin{center}
%%%%% SEAIR diagram
\resizebox{!}{6cm}{
\begin{tikzpicture}[
  square/.style={rectangle, draw=black, minimum width=0.8cm, minimum height=0.8cm, rounded corners=.1cm, fill=blue!8},
  travel/.style={circle, draw=black, minimum width=0.85cm, minimum height=0.8cm, fill=green!8},
  report/.style={shape=regular polygon, regular polygon sides=8, draw, fill=red!8,minimum size=0.9cm,inner sep=0cm},
  bendy/.style={bend left=25},
  >/.style={shorten >=0.25mm}, % redefine arrow to stop short of node
  >/.tip={Stealth[length=1.5mm,width=1.5mm]} % redefine arrow style
]
\tikzset{>={}}; % this is needed to implement the arrow redefinition
\tikzset{every path/.style={line width=0.8 pt}}

\draw[rounded corners, green, thick] (-1.2, 1) rectangle (5.2, 2.5) {};
\draw[rounded corners, blue, thick] (-1.2, -2.4) rectangle (7.7, 0.7) {};
\draw[rounded corners, red, thick] (3.75, -4.15) rectangle (10.2, -2.85) {};
\node (between) at (5.4,2.3) [green, anchor=north west, font=\large] {Travel between cities};
\node (within) at (7.9,0.5) [blue, anchor=north west, font=\large] {Disease dynamics within cities};
\node (report) at (10.4,-3.05) [red, anchor=north west, font=\large] {Reported cases};

@@

For our metapopulation system, the sub-populations are 373 provincial cities in China (meaning cities with administrative responsibility for an entire region) and the data are daily reported COVID-19 cases. 
COVID-19 dynamics are represented by a Susceptible Exposed Asymptomatic Infectious Recovered (SEAIR) epidemic model. 
Questions of urgent interest early in the pandemic include the relative transmissibility of reported to unreported cases, the fraction of unreported cases, and the effect on transmission of movement restrictions imposed on and around January 23 \cite{li20}.
The model structure is illustrated by the flow diagram in Fig.~\ref{fig:flow_diagram}.
The Methods section provides additional description, with Eq.~(\ref{eq:muSE}) giving the modeled rate at which susceptible individuals become infected.
The complete model specification and estimated parameter values are in Supplementary Sec.~{\suppSecOurModel}.

We consider different model implementations within this structure.
Our starting point is model {\LiMobility} which is based on the model of \cite{li20} and is described in Supplementary Sec.~{\suppSecLiModel}.
We consider the full dataset, from January 10 to February 8, with transmission parameters re-estimated following the lockdown on January 23; these correspond to the periods~1 (January 10 to January 22) and 3 (January 24 to February 8) of \cite{li20}.
Some minor differences between {\LiMobility} and \citep{li20} were introduced to enable us to place their model within the general metapopulation framework of spatiotemporal partially observed  continuous-time Markov processs described by \citep{ionides23-jasa}.
Despite these modifications, simulations from {\LiMobility}, using the parameters of \citep{li20}, closely match simulations from the code provided by \cite{li20} (Supplementary Sec.~\suppSecLiModel).
However, inspection of the mobility data reveals that some small cities have no recorded incoming travelers, and therefore no possibility of a SARS-CoV2 introduction within {\LiMobility} (or the model of \cite{li20}) (Supplementary Sec.~\suppSecMobility).
This minor limitation formally results in a likelihood of zero for {\LiMobility} (i.e., it is impossible for the simulation model to reproduce the observed spatiotemporal dataset), and hence a log-likelihood of $-\infty$.

@@

We addressed the problematic mobility data in {\LiMobility} by adding some additional transportation based on a gravity movement model, as described in Supplementary Sec.~{\suppSecMobility}, giving rise to model~{\LiParams}.
We implemented an additional adjustment between models {\LiMobility} and {\LiParams} to align the measurement model with the ensemble Kalman filter (EnKF) inference method presented by \cite{li20}.
That EnKF implementation involved specifying a quantity called the observation error variance, defined as a function of the observed cases, to quantify the uncertainty in the measurements.
Within the POMP specification, the measurement variance can depend on the latent state but not directly on the observed data.
To interpret the choice of EnKF observation variance within the POMP framework, we specified the measurement model for~{\LiParams} to have equivalent scaling to the choice of \citep{li20}, but with dependence on the reported cases replaced by dependence on the modeled, but unobserved, exact case count.

@@

Table~1 shows that model~{\RevisedModelUnconstrained} outperforms simple statistical benchmarks, obtaining a competitive likelihood with relatively few parameters.
From a statistical perspective, {\RevisedModelUnconstrained} is therefore an adequate statistical description of the data.
However, some parameters of {\RevisedModelUnconstrained} were weakly identified by the data, especially in the pre-lockdown time interval within which there were relatively few reported cases (Supplementary Sec.~\suppSecUnconstrained).
When the evidence about the model parameters in the data is weak, the ambiguity may be resolved by other, unmodeled and poorly understood, aspects of the data.
This risks leading to undesirable situations where substantial conclusions about questions of interest could be driven by the weaknesses of the model rather than its strengths.
In Supplementary Sec.~{\suppSecUnconstrained}, we show how the flexibility of {\RevisedModelUnconstrained} can be used to obtain a high likelihood via an unplausibly long estimated duration of infection during the pre-lockdown period, with the estimate suddenly reducing after lockdown.
 We resolved this issue by constraining the latent and infectious periods to be the same before and after lockdown, leading to model {\RevisedModelConstrained}.
 The additional constraints of {\RevisedModelConstrained} lead to a small loss of likelihood compared to {\RevisedModelUnconstrained}, but the fit remains competitive compared to the benchmark models, and the stronger identifiability facilitates the interpretation of estimated parameters.
Calculating the log-likelihood for each model in Table~\ref{tab:loglik} requires extensive computation to produce a single number which contains essentially all the information about the statistical fit of the model.
However, deeper investigation is required to understand what characteristics of the models and data causes the differences in these numbers, and the practical consequences of the numerical results.
As a starting point, Fig.~\ref{fig:panel_plot} plots the data next to simulations from models {\LiMobility} and {\RevisedModelConstrained}.
Visually, the comparison confirms {\RevisedModelConstrained} as a reasonable representation of the data.
Both {\LiMobility} and {\RevisedModelConstrained} overestimate cases before day 14, but the context of rapidly increasing awareness and growing diagnostic capabilities is hard to quantify.
